---
title: "Basic GLM with booleans template"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03_template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# STEP 1: Server connection

Change the credentials to your own:

```{r user_credentials}
your_url <- "https://opal.isglobal.org/brge"
your_user <- "administrator"
your_password <- "password"
your_resource <- "omop_demo.tufts"
```

Then run the following code to connect to the server and create a database interface:

```{r server_connection, message=FALSE, warning=FALSE}
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)
library(dsOMOPHelper)

builder <- newDSLoginBuilder()

builder$append(server="server1", 
               url=your_url,
               user=your_user, 
               password=your_password, 
               driver = "OpalDriver")

logindata <- builder$build()

conns <- datashield.login(logins=logindata)

o <- ds.omop.helper(
    connections = conns,
    resource = your_resource, 
    symbol = "original_dataset"
)
```

# STEP 2: Dataset exploration

We will first retrieve the concept catalogs for conditions and observations:

```{r dataset_exploration}
condition_list <- o$concepts("condition_occurrence", .Machine$integer.max)$server1
observation_list <- o$concepts("observation", .Machine$integer.max)$server1
```

Now you can explore the available concepts and decide which ones you want to use in your analysis.

We strongly recommend saving these lists to CSV files and exploring them in a comfortable environment. This will allow you to identify the exact name of the variables you want to use in your analysis:

```{r write_csv, eval=FALSE}
write.csv(condition_list, file = "condition_list.csv", row.names = FALSE)
write.csv(observation_list, file = "observation_list.csv", row.names = FALSE)
```

Otherwise, you can check the concept catalogs directly by printing them:

```{r head_concept_catalogs, echo=FALSE}
# Keep only the first 5 rows of condition_list for the demo
condition_list <- head(condition_list, 5)
# Keep only the first 5 rows of observation_list for the demo
observation_list <- head(observation_list, 5)
```
```{r print_concept_catalogs}
print(condition_list)
print(observation_list)
```

(The actual concept list will be much longer, this is just for demonstration purposes)

# STEP 3: Dataset construction

## Variable definition

Change the following parameters to fit your needs (you will need to use the concept IDs from the `condition_list` and `observation_list`):

Define the outcome variable (from the `condition_list`):

```{r define_variables_outcome}
outcome_concept_id <- 255573 # Chronic Obstructive Pulmonary Disease (COPD)
```

Define the predictor variables (conditions from the `condition_list` and observations from the `observation_list`):

```{r define_variables_predictors}
observation_list <- c(
    4005823 # Tobacco use
)

condition_list <- c(
    317009, # History of asthma
    436070, # Rheumatoid arthritis
    80809   # Vitamin D deficiency
)
```

From this point, the script will perform the rest of the process automatically. You can tweak the previous settings to fit your needs. 

# Automatic dataset construction

The script will perform the following steps automatically:

1. Retrieve the variables from the database:

```{r data_retrieval}
o$auto(
    table = "condition_occurrence", 
    concepts = c(outcome_concept_id, condition_list),
    columns = c("condition_occurrence_id") 
    # We only want the condition occurrence ID to act as a boolean indicating 
    # the presence of the condition occurrence
)

o$auto(
    table = "observation", 
    concepts = observation_list,
    columns = c("observation_id")
    # We only want the observation ID to act as a boolean indicating 
    # the presence of the observation
)
```

2. Match the concept IDs to the concept catalogs:

```{r custom_make_names}
custom_make_names <- function(name) {
  name <- make.names(name)
  name <- tolower(name)
  name <- gsub("\\.", "_", name)
  name <- gsub("_+", "_", name)
  name <- gsub("^_|_$", "", name)
  return(name)
}

# Retrieve the concept catalogs for condition occurrence and observation
all_concepts_condition <- o$concepts("condition_occurrence", .Machine$integer.max)$server1
all_concepts_observation <- o$concepts("observation", .Machine$integer.max)$server1

# Match the concept IDs to the concept catalogs
matched_concept_condition <- all_concepts_condition[all_concepts_condition$concept_id == outcome_concept_id, ]
matched_concepts_conditions <- all_concepts_condition[all_concepts_condition$concept_id %in% condition_list, ]
matched_concepts_observation <- all_concepts_observation[all_concepts_observation$concept_id %in% observation_list, ]
```

3. Convert the variables to boolean:

```{r convert_to_numeric_and_boolean}
convert_to_numeric_and_boolean <- function(variable_name, id_type, conns) {
  # Construct the full variable name in the format "original_dataset$variable_name.id_type"
  full_variable_name <- paste0("original_dataset$", variable_name, ".", id_type)
  # Create a new variable name for the numeric conversion
  new_numeric_name <- paste0(variable_name, "_numeric")
  
  # Convert the original variable to numeric
  ds.asNumeric(
    x.name = full_variable_name, 
    newobj = new_numeric_name, 
    datasources = conns
  )

  # Convert the numeric variable to boolean
  # True (1) if not equal to 0, False (0) otherwise
  # NA values are assigned 0
  ds.Boole(
    V1 = new_numeric_name, 
    V2 = 0, 
    Boolean.operator = "!=", 
    numeric.output = TRUE, 
    na.assign = 0, 
    newobj = variable_name
  )
}

# Get the name of the outcome variable as it will be in the dataset
outcome_variable <- custom_make_names(matched_concept_condition$concept_name)
# Perform the type conversion of the outcome variable to numeric and then to boolean
convert_to_numeric_and_boolean(outcome_variable, "condition_occurrence_id", conns)

# Loop through every recognized condition occurrence concept
for (concept in matched_concepts_conditions$concept_name) {
  # Get the exact name of the condition occurrence as it will be in the dataset
  variable_name <- custom_make_names(concept)
  
  # Perform the type conversion of the condition occurrence to numeric and then to boolean
  convert_to_numeric_and_boolean(variable_name, "condition_occurrence_id", conns)
}

# Loop through every recognized observation concept
for (concept in matched_concepts_observation$concept_name) {
  # Get the exact name of the observation as it will be in the dataset
  variable_name <- custom_make_names(concept)
  
  # Perform the type conversion of the observation to numeric and then to boolean
  convert_to_numeric_and_boolean(variable_name, "observation_id", conns)
}
```

4. Create a new table with the prepared outcome and predictors:

```{r create_glm_table}
# Get the exact name of the variables as they will be in the dataset
variable_names <- c(
  unname(sapply(matched_concepts_observation$concept_name, custom_make_names)),
  unname(sapply(matched_concepts_conditions$concept_name, custom_make_names)),
  custom_make_names(outcome_variable)
)

# Create a new table with the prepared outcome and predictors
ds.cbind(
  x = variable_names,
  DataSHIELD.checks = FALSE,
  newobj = "glm_table",
  datasources = conns
)
```

# Generalized Linear Model

It then runs the GLM with the prepared dataset:

```{r glm, warning=FALSE}
# Define the formula for the GLM
formula <- paste0(
  "glm_table$", outcome_variable, " ~ ", 
  paste(
    c(
      paste0("glm_table$", custom_make_names(matched_concepts_observation$concept_name)),
      paste0("glm_table$", custom_make_names(matched_concepts_conditions$concept_name))
    ),
    collapse = " + "
  )
)

# Fit the GLM
ds.glm(
  formula = formula, 
  data = "glm_table", 
  family = "binomial"
)
```
